Imputation of Incomplete Multilevel Data with R

### 2024-01-16 Manuscript/manuscript.qmd
Suggestions for background section

There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.

@@

Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009). The last restriction is also required when imputations are performed on each cluster separately.  When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster. However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e. the number of clusters is adequate. (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008). Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020). For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions. (Zhang,2023).

@@

### Focus on the Main Analysis

When dealing with incomplete clustered data, start by looking at your research questions and planned analysis. Pretend there are no missing data at first. This will help you understand your research hypotheses, the main statistical model, and give you insights into your data's structure (refer to the level table), variable types (e.g., confounders, auxiliary variables), and other considerations such as variable relationships, interactions or polynomial terms.

@@

Now, explore what's in your data. Use simple tools like histograms and QQ plots to understand variable distributions, plausible range of values and identify errors or outliers. Check interactions between predictors with scatter plots, look for collinearity with VIF or correlations plots, this may also help you to identify non-considered relationships that may affect the main model.  You might also consider to test assumptions related to your response variable such as variance homogeneity, independence between observations (eg. ACF, variagrams).  This will help you choose an imputation model that suits your data. In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between the 2l and 1l methods for imputation.

Next, explore the missingness. Look at the \textbf{proportion of missing values} in the dataset variables. This helps find potential predictors and cut down on unnecessary variables in the imputation model. Doing this can lower the risk of multicollinearity or computational issues, especially with certain parametric imputation methods. You can also identify predictors for the imputation model by using inflow criteria to see connections between missing data in one variable and observed variables, and outflow criteria to identify connections between observed values in one variable and missing data in others.

Check the missing patterns at cluster level, this can help you to select the most appropiated imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).

@@

### Pre-imputation
Clean up your dataset before the imputation process. Keep initially the essential variables for your main model, but include extra variables if needed for specific procedures during analysis (e.g. confounders on balancing procedures). Think about adding other useful variables, even if they're not in the main model, such as instrumental or auxiliary variables which might boost your parameter estimates, especially if they're associated to the probability of missingness for some incomplete variables.

Figure out if you can directly impute incomplete variables by just use deductive imputation. This involves inferring missing values based on logicallogical connections between variables. It's especially handy for variables that depend on each other, like calculating BMI from weight and height.

Deductive imputation is also useful for getting values for level-1 variables from level-2 ones, like in Individual Participant Data (IPD). In these situations, you can guess missing information from metadata or by using or through deduction based on time or protocols. For example, you could deduce missing test values for patients who have passed away.

>>>>

There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of a cluster indicator variable, performing a separate imputation process for each cluster, or performing a imputation for each cluster simultaneously by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.

@@

<!-- I don't understand the following paragraph. What is meant with "is restricted in datasets..."? -->
Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009). The last restriction is also required when imputations are performed on each cluster separately.  When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster. However, this strategy is best suited for balanced data (Grund, 2017) and when the number of clusters is adequate sufficient for the random effects model (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main model, for instance the cluster indicator strategy bias the estimator when the model is based on a hierarchical model (Taaljard,2008). Although an imputation strategy congenial with the main model is preferred, it is important to consider whether a complex imputation model is appropriate for the data, as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios (Bailey 2020). For instance, in causal effect analysis, separately imputating data sets may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions. (Zhang,2023).
<!-- What is an exposure cluster? -->

@@

@@

First, explore the data, preferably using visual tools. This will help you choose an imputation model that suits your data. In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between methods for dependent (such as clustering) or independent data.

Next, explore the missingness. Look at the \textbf{proportion of missing values} in the dataset variables. Variables for which most data is missing may contribute little in the analysis, and may be removed from the analysis model. Removing them also from the imputation model can lower the risk of multicollinearity or computational issues, especially with certain parametric imputation methods. You can also identify predictors for the imputation model by using inflow criteria to see connections between missing data in one variable and observed variables, and outflow criteria to identify connections between observed values in one variable and missing data in others.

<!-- What are inflow and outflow criteria? -->

Check the missing patterns at cluster level, this can help you to select the most appropriate imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns).

@@

<!-- What does it mean if an estimation model is not able to handle missing data? What do you mean with estimation model? -->
<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results. The text appears to suggest that in that case ML based methods are better. -->

### Pre-imputation
Clean up your dataset before the imputation process. For imputation, include the variables necessary for your analysis, depending on your analysis this may be the exposure or intervention, outcome, covariates, confounders, stratification factors, and the cluster variable. Think about adding other useful variables, even if they're not in the main model, such as instrumental or auxiliary variables which might improve the accuracy of the imputation model, especially if they're associated to the probability of missingness for some incomplete variables.

Figure out if you can directly impute incomplete variables by using deductive imputation. This involves inferring missing values based on logical connections between variables. It's especially useful for variables that depend on each other, like calculating BMI from weight and height.

Deductive imputation is also useful for getting values for level-1 variables from level-2 ones, like in Individual Participant Data (IPD). In some situations, you can derive missing information from metadata or protocols. For example, you could deduce missing test values for patients who have passed away. 
<!-- I don't think this is the best example. Usually, test scores after death are not of interest. The test score before death may be missing for a reason, but this should be derived based on clinical knowledge. The example is too vague (and I can't readily improve it.) -->
